Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-19 forecastflows.tex
Small edits to current draft with some call-out notes

\documentclass[11pt]{article}
\usepackage[top=1.00in, bottom=1.0in, left=1in, right=1in]{geometry}
\renewcommand{\baselinestretch}{1.1}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{multicol}
\usepackage{tcolorbox}
\usepackage{caption}

@@

\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology } 
\date{\today}
\author{Victor Van der Meersch, J. Regetz, T. J. Davies \& EM Wolkovich}
\maketitle
% $^*$ Says he is happy to help and give friendly review, but not sure he will reach level of co-author (though Lizzie used his LPI example).

{\bf Deadline:} 1 May 2025 (was 1 April 2025)

@@

@@

We argue that current debates and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, forecasting-focused process-based models often develop by adding new separate layers or components. These new parts are often disconnected from the original research aim, its data stream and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major degeneracies.

Workflows that fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections, could reduce many of these problems. In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to undertand when, where and why different models diverge. Towards this aim,  we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.

@@

Quantitative science relies on a model-based framework, to confront hypothesis with data \citep{Chamberlin:1965cd}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses---following the principles of pre-registration, where hypotheses and methodology are defined prior to data collection.  % jdavies15april: could refer to the 'pre-registration' movement here ...
But this idealized method often does not apply to the reality of ecological research. Many important questions cannot be addressed through controlled experiments and replications. Instead, we must often rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}, and most of macroecological insights emerge from exploring patterns in these datasets (exploratory data analysis).

This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in term of data complexity and societal needs---instead may lead to persistent problems. % jdavies15april: what challenges? forecasting from imperfect data or the urgent need to generate forecasts?
Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way % jdavies15april: not obvious what this means... [FLAG]
that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig. \ref{fig:modeldata}). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (Fig. \ref{fig:modeldata}). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole. % jdavies15april: is this particularly specific to process based models? 
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow. 
% mention the disconnection between trends and forecasts?

@@

We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation % jregetz16april: Maybe “data simulation” needs to appear in the figure? Or wait to mention simulation until referencing Fig 2? [FLAG: what do you think? I agree you should choose on of the two options]
(Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, fig. \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This step should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error. % vvdm16april: to answer Jonathan comment about verbal model
From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, fig. \ref{fig:workflow}). The general idea is to start with a relatively simple model that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires to think about alternative parameterizations).

With a model in place the next step focuses on testing and understanding it via data simulation (step 3, fig. \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data. % This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model. % jdavies15april: I think you could introduce this as a separate step as it seems there is a tendency for people to want to do this immediately (i.e. simulate fake data that looks like their data - I am looking at Max and Isidora ...), but I don't think this is helpful when first structuring a model. [FLAG: yes I can understand this point...]
% For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. [FLAG: so I moved this sentence below, when we say "Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data"
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, fig. \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question, or re-evaluate the model of how we think the system works. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. This leads to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, fig. \ref{fig:workflow}). This retrodictive check allows model output to be compared to observations. % jregetz16april: do we need to define “retrodictive”?
% The workflow encourages a focus on the full model, and replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting. % jregetz16april: Not sure what this means? jdavies did not like it either! (he thought it was a typo), vvdm16april: reformulate:
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting. 
Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data. The adoption of the workflow in macroecological studies---where model building is often informed by patterns in the data---would also make the exploratory analysis more transparent (as an explicit preliminary step for entering the workflow) and would compel researchers to come out with a research question.

@@

% More details, what's special! Feedbacks!
A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
Further, this step might reveal that some parameters are highly non-identifiable,
flagging the need to change the model structure---before incorporating real observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies may indicate a missing key driver---perhaps an expected outcome if we known our initial model was too simplistic. We can refine the model to integrate the missing process and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

@@

%jregetz16april: One thing I’m stuck on (and could’ve written this note earlier too). Is “trends” really essential? Or is this just shorthand for “simple, one/few parameter, non-mechanistic forecasting models”? If yes, can we generalize that in the MS overall, or at least say so up front and then say e.g. “we’ll refer to these as trend models for brevity”? If no, and something about trend estimates is specifically key to the paper, can we say more about *that*? [FLAG]

Ecologists today have amassed data on populations and species across the globe; they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction of population and biodiversity metrics \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}. While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be much reduced and more rapidly resolved through use of an improved workflow. % Further, an improved workflow could give more rapid and coherent estimates, which could make it easier for  policy-makers to develop and advance initiatives aimed at slowing declines.

An improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make what now appear as major discrepancies more obviously shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model. We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models. While ecologists have started to use simulated data more to understand potential limitations of their models and data combined, this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility.
% jdavies15april: but there is a history of using null models (e.g. Gotelli &co) that is perhaps not so different - here the 'simulated' data is some sampled process of the observed data, but we can estimate standard effect sizes that are part of our statistical model. I know this is different to the simulations you propose based on some generative process, but ecologists should be able to able to make this additional step.

@@

\subsection{Forecasting}

Ecological forecasting is a broad field with a diverse range of methods. Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond. Newer models generally incorporate greater complexity and an ever-growing number of parameters, making it more difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.

Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome. In climate modeling, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. % jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting. [FLAG: agree but not really relevant to PBM? Maybe it's not clear enough we are focusing on PBM in this section...]
If a model becomes too complex, it may become a black box, % mention black box here to line up with the beginning of next paragraph
and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}. % jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty .... [FLAG: agree but not really relevant to PBM? Maybe it's not clear enough we are focusing on PBM in this section...]

% jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties? This could help illustrate potential degeneracies etc.
% [FLAG: I tried something, but I'm not really illustrating degeneracies + I don't like citing myself as much + I'm not really satisfied]
An illustration of the issues with the current forecast workflow is the process-based models used to project species or ecosystem distributions. Most of the focus is on the model projections---often represented as maps without uncertainties. Model evaluation generally focuses on how well these projections match observed distributions, sometimes under different climatic conditions to challenge the models \citep{VanderMeersch2025a}---most of the time evaluating only of the many different output variables of the model. The model building and calibration steps are usually limited to the Methods section, and estimated parameter values are rarely discussed \citep[e.g.][]{Conradi2024, VanderMeersch2025a}. The complexity of these models often makes it difficult to assess the robustness of their calibration, whether for a reviewer or a reader. If ever done, the assesment of the model degeneracies is often done as a a separate work, exposing model degeneracies later on \citep{VanderMeersch2025b}, and not required as a preliminary step.

Applying the workflow to process-based models would help open the black box.
Each successive step of model development in the workflow may highlight current problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two and help expose potential degeneracies in the model design. %jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below) [FLAG: agree, and made me realize it was not 100% clear in my mind. As I understand it, non-identifiability concerns one parameter, and degeneracy concerns the all model?... but not sure]
Uncovering degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight degeneracies by requiring more explicit model calibration (i.e. data fitting), % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model (where certain parameters cannot be uniquely estimated). Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.

% jdavies15april: might be better grouped together in concluding section
% [FLAG: I feel if we want to put this in the concluding section, we would need to talk also about the "opportunities to bridge statistical and process-based", rather than keeping it in the box? So, what should we put in the box?]
Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them. The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model. It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.  % or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza! disease! pathogens!)... or "rather than constantly expanding models by adding ever more processes"? 
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty. Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification.

This shift would present a significant challenge---as it would likely reveal many issues related to model degeneracies and data limitations before achieving robust inference. But ultimately, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support.

@@

In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data. 
The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these conflicting results. % jdavies16april: could highlight how influential the LPI has become (recent reports suggest a 73% ave. pop decline since 1970. This would be massive!
With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach. For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}. 
% jdavies16april: One of my major issues with all these studies is that they have lost track of the purpose of the index, which is to capture general trends in animal populations. We do not actually care (much) about the population trends in the LPI, but how these are representative of global trends. For this we need to predict outside of the set ... [FLAG]

>>>>

\documentclass[11pt]{article}
\usepackage[top=1.00in, bottom=1.0in, left=1in, right=1in]{geometry}
\renewcommand{\baselinestretch}{1.1}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{multicol}
\usepackage{tcolorbox}
\usepackage{caption}
\usepackage{todonotes}

@@

\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology } 
\date{\today}
\author{Victor Van der Meersch, J. Regetz, T. J. Davies \& EM Wolkovich}
\maketitle

{\bf Deadline:} 1-ish May 2025 (was 1 April 2025, getting in by May 8th would be good!)

@@

% Jim, if you want to add any of this into the abstract, please do!
%There's a fundamental problem in how a lot of empirically but observationally driven (i.e., non-experimental) research is done in ecology and related fields. The crux of the issue comes down to a couple of things. First, the research community has two very distinct paradigms for modeling the system or phenomenon or interest: either fitting simple trend-like models with only one or a few parameters, or developing forecast models that often include potentially complex, mechanistic, and/or black box submodels largely unconstrained by data. Second, regardless of the approach, these models tend to be developed not only in isolation of one another, but also without a coherent strategy for linking modeling/statistical decisions, data, and scientific understanding (both as input and output) throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development, incorporating sound statistical and scientific practices, and especially relying on data simulation to inform decisions at multiple steps in the process.

@@

We argue that current debates and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, forecasting-focused process-based models often develop by adding new separate layers or components. These new parts are often disconnected from the original research aim, its data stream and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below).

Workflows that fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and potential problems, before estimating its parameters and making projections, could reduce many of these problems. In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where and why different models diverge. Towards this aim,  we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.

@@

Quantitative science relies on a model-based framework, to confront hypothesis with data \citep{Chamberlin:1965cd}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses. This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection. %add cites for prereg?
But this idealized method often does not apply to the reality of ecological research. Many important questions cannot be addressed through controlled experiments and replications. Instead, we must often rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}, and most of macroecological insights emerge from exploring patterns in these datasets (exploratory data analysis).

This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in term of data complexity and societal needs---instead may lead to persistent problems. % jdavies15april: what challenges? forecasting from imperfect data or the urgent need to generate forecasts?
Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way---without the checks and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (Fig. \ref{fig:modeldata}). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (Fig. \ref{fig:modeldata}). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole. % jdavies15april: is this particularly specific to process based models? 
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow. 
% mention the disconnection between trends and forecasts?

@@

We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation % jregetz16april: Maybe “data simulation” needs to appear in the figure? Or wait to mention simulation until referencing Fig 2? [FLAG: what do you think? I agree you should choose on of the two options]
(Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, fig. \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This step should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error. % vvdm16april: to answer Jonathan comment about verbal model
From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, fig. \ref{fig:workflow}). The general idea is to start with a relatively simple model that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterizations).

With a model in place the next step focuses on testing and understanding it via data simulation (step 3, fig. \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data. % This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model. % jdavies15april: I think you could introduce this as a separate step as it seems there is a tendency for people to want to do this immediately (i.e. simulate fake data that looks like their data - I am looking at Max and Isidora ...), but I don't think this is helpful when first structuring a model. [FLAG: yes I can understand this point...]
% For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. [FLAG: so I moved this sentence below, when we say "Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data"
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. At this stage the focus is understanding the model, so researchers may have to spend a while to think through what parameters are reasonable or not. This is also a step for making sure the model is working as we expect, so simulated data often would have low error and high sample sizes (to make sure the model returns the parameter values used to generate the simulated data),

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, fig. \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations. 
%emwApr19: I commented out the below as I think linking difficulty in fitting the model to problems with the data is pretty tricky (difficulty in fitting the model can point to many things!) and thus too confusing to include here. 
% Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question, or re-evaluate the model of how we think the system works. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. 
These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, fig. \ref{fig:workflow}). This retrodictive check allows model output to be compared to observations. % jregetz16april: do we need to define “retrodictive”?
% The workflow encourages a focus on the full model, and replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting. % jregetz16april: Not sure what this means? jdavies did not like it either! (he thought it was a typo), vvdm16april: reformulate:
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting. 
Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data. The adoption of the workflow in macroecological studies---where model building is often informed by patterns in the data---would also make the exploratory analysis more transparent (as an explicit preliminary step for entering the workflow) and would compel researchers to more clearly develop a research question before extensive model fitting.

@@

% More details, what's special! Feedbacks!
A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
flagging the need to change the model structure---before incorporating real observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es) (if we can identify them) and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

@@

%jregetz16april: One thing I’m stuck on (and could’ve written this note earlier too). Is “trends” really essential? Or is this just shorthand for “simple, one/few parameter, non-mechanistic forecasting models”? If yes, can we generalize that in the MS overall, or at least say so up front and then say e.g. “we’ll refer to these as trend models for brevity”? If no, and something about trend estimates is specifically key to the paper, can we say more about *that*? [FLAG]
%emwApr19: I think the only key part is that biodiversity trends and trends over time are big in forecasting ... as long as we don't lose that I think it would be great to make the suggested change -- Jim, can you see if you can edit it on the next round?

Ecologists today have amassed data on populations and species across the globe; they have also engaged in an increasing number of debates on regional and global trends over time, with arguments over the magnitude and even direction of population and biodiversity metrics \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}. While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be much reduced and more rapidly resolved through use of an improved workflow. % Further, an improved workflow could give more rapid and coherent estimates, which could make it easier for  policy-makers to develop and advance initiatives aimed at slowing declines.

An improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make what now appear as major discrepancies more obviously shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model. We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models. This step builds somewhat on the skills needed for null models \citep{Gotelli:2012oz}, but with a shift in focus towards the specific bespoke model at hand. Ecologists have started to use simulated data more to understand potential limitations of their models and data combined, but this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility if used in an iterative workflow.

@@

\subsection{Forecasting from process-based models}

Ecological forecasting is a broad field with a diverse range of methods. Our second case-study focuses on process-based modeling, which is often considered the gold standard for forecasting in ecology \citep{Urban2016, Pilowsky2022} and beyond. Newer models generally incorporate greater complexity and an ever-growing number of parameters, making it more difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.

Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome. In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. \todo[size=\tiny]{jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting. Reply: we agree but not really relevant to PBM? Maybe it's not clear enough we are focusing on PBM in this section, which we tried to change above, see if it works}
If a model becomes too complex, it may become a black box, % mention black box here to line up with the beginning of next paragraph
and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}. \todo[size=\tiny]{jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty .... Reply: Same as just above.}

Process-based models used to project species or ecosystem distributions highlight some of these problems. Most of the focus is on the model projections---often represented as maps without uncertainties. Model evaluation generally focuses on how well these projections match observed distributions, sometimes under different climatic conditions to challenge the models \citep{VanderMeersch2025a}---most of the time evaluating only one of the many different output variables of the model. %The model building and calibration steps are usually limited to the Methods section, and estimated parameter values are rarely discussed \citep[e.g.][]{Conradi2024, VanderMeersch2025a}. 
The complexity of these models often makes it difficult to assess how well the potential parameter space was searched, and whether there was any potential non-identifiability---such work is generally done as a separate effort, exposing potential model problems later on \citep{VanderMeersch2025b}, and not required as a preliminary step. \todo[inline, size=\tiny]{For above, addressing: jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties? This could help illustrate potential degeneracies etc. Reply: Tried something, but not sure it works so feedback much appreciated!}

%emw19Apr -- Can we just get away with using non-identifiable and save the reader also needing to digest degenerate? (https://betanalpha.github.io/assets/case_studies/identifiability.html)? I tried that on this draft, see how it works!
Applying the workflow to process-based models would help open the black box.
Each successive step of model development in the workflow may highlight current problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two and help expose potential parameter identifiability issues in the model design. %jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below) 
Uncovering identifiability issues would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight model problems by requiring more explicit model calibration (i.e. data fitting), % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model. Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.

\todo[inline, size=\tiny]{For below: jdavies15april: might be better grouped together in concluding section. Reply: if we want to put this in the concluding section, we would need to talk also about the "opportunities to bridge statistical and process-based", rather than keeping it in the box? So, what should we put in the box?}
Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them. The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model. It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.  % or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza! disease! pathogens!)... or "rather than constantly expanding models by adding ever more processes"? 
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty. Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification.

This shift would present a significant challenge---as it would likely reveal many issues related to non-identifiability in models and data limitations before achieving robust inference. But ultimately, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support.

@@

In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data. 
The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these conflicting results. % \todo[]{could highlight how influential the LPI has become (recent reports suggest a 73\% ave. pop decline since 1970. This would be massive!}
With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach. For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}. 
% \todo[]{davies16april: One of my major issues with all these studies is that they have lost track of the purpose of the index, which is to capture general trends in animal populations. We do not actually care (much) about the population trends in the LPI, but how these are representative of global trends. For this we need to predict outside of the set}
